Using VariantSpark to learn population structure from genomic profiles

### 2015-10-27 doc/publications/Recomb15/recomb15.tex
updates

\maketitle
%
\begin{abstract}
Using our 
We apply various machine learning algorithms to the 1092 individuals in the 1000 Genomes Project dataset.
Although each algorithm we investigate can cluster individuals from three populations with near perfect accuracy,
we find logistic regression to be superior. Comparing  With the entire dataset,

@@

Using \kMeans{} we observed a perfect separation for the super populations African (AFR), East Asian (EAS) and European
(EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR
individual form an independent cluster while the majority falls into the EUR cluster.  We quantified this by calculating
the \ARI\ ( ARI), which compares the predicted and annotated labels. \kMeans achieved an ARI of 0.84, with -1 being
random and 1 a perfect clustering.  We argue that the suboptimal clustering is a result of using an unsupervised method
rather than an insufficient number of genomic markers~\cite{Patterson2006} and hence hypothesise that a supervised
machine learning method may be capable of separating these subgroups of low genetic divergence.  Indeed, bigdata
genomics recently investigated the supervised method, deep learning, for clustering the 1000 genomes data by
ethnicity~\cite{Ferguson}.  \marginpar{``bigdata genomics'' seems to be a domain name \url{http://bdgenomics.org/} --
  how should we cite this}

In this paper we therefore systematically evaluate supervised and unsupervised methods for determining population
association and compare our results to the ones obtained using deep learning.
%\marginpar{TODO Rob can you think of a better description of what it does}
One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary
by fitting a linear combination of the feature variables.  In this appliction it is not fesible to model interactions,
so the model is retricted to main effects.

%\marginpar{TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}.}  
We also investigate more complex supervised
approaches, namely Random Forest (RF), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study
around genetic susceptibility to obesity that only five of the 97 genomic loci have an independent association signal
the rest seem to jointly predict the complex metabolic phenotype.  We therefore investigate whether ethnicity is a
similarly complex phenotype that requires higher order dependencies between features to be taken into
account. %, as done by RF and DL.

RF inherit many of the benefits of decision trees. It handles large numbers of variables transparently and
provides an unbiased estimate of the generalization error (in addition, it handles missing values and unbalanced
population classes sizes). They provide a measure of variable importance and allow for interactions between
variables to be investigated. However, in this problem, with many millions of variables, inspecting plots of interactions
is not feasible.  We note that decision trees (and random forests) model local variable interactions -- there are 
no main effects or global interaction effects.

@@

However, training a supervised learning method on such an expansive dataset as the 1000 genomes is computationally 
challenging and traditional massively parallel approaches to data processing are not easily applicable to machine learning 
tasks, as they iteratively refine models based on information from the full dataset.

@@

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.  
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to a limitation of the Java virtual machine (maximum array size) that causes {\sc Mllib}'s
random forest implementation to fail. This error is due to the number of features present in each sample,
rather than the number of samples, and a potential solution is to either limit the features through feature selection, or to simply use a subset of the dataset.
We therefore use only chromosome 22 (494,328 variants) for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes Project dataset using the annotated labels of all
super-populations.  We perform 10-fold cross-validation and calculate the Adjusted Rand Index (ARI) for use as a metric for model quality.
The ARI allows us to compare the classification results with \kMeans{} clustering results, as it takes the partitioning of the dataset into account, rather than just labels.
report the mean error over all training and testing folds
along with the standard error of the mean (SEM)
We also calculate the error
for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.
We perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean ARI of the
resulting models along with the standard error.

As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately, with an average ARI over ten models of \(0.596 \pm 0.067\), see table~\ref{RF}.
%Although capable of fitting a relatively accurate model (\(ARI=0.837\)), The average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\), however it is outperformed by the LR, with an ARI of \(0.923 \pm 0.015\).
This indicates that while applying a supervised learning method results in higher accuracy, a multivariate approach may not be needed to accurately fit ethnicity.

@@

\subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
In this section we investigate whether another multivariate approach, DL, is able to achieve a better result than
RF by again comparing it against \kMeans{} and logistic regression.  We compare against the reported accuracy of the
deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}.
As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW
(Americans of African Ancestry in SW USA), CHB (Han Chinese in Beijing, China), and GBR (British in England and
Scotland).

>>>>

\maketitle
%
\begin{abstract}
Using VariantSpark, we compare various machine learning algorithms on the 1092 individuals in the 1000 Genomes Project dataset.
Although each algorithm that we investigate can classify or cluster individuals from three populations with near perfect accuracy,
we find logistic regression to outperform the others. We observe a similar outcome on the entire dataset, with logistic regression
classifying the majority of the individuals correctly.

@@

Using \kMeans{} we observed a perfect separation for the super populations African (AFR), East Asian (EAS) and European
(EUR), however when adding the fourth super-population group, American (AMR), we observe that only a subset of the AMR
individual form an independent cluster while the majority falls into the EUR cluster.  We quantified this by calculating
the \ARI\ (ARI), which compares the predicted and annotated labels. \kMeans achieved an ARI of 0.84, with -1 being
random and 1 a perfect clustering.  We argue that the suboptimal clustering is a result of using an unsupervised method
rather than an insufficient number of genomic markers~\cite{Patterson2006} and hence hypothesise that a supervised
machine learning method may be capable of separating these subgroups of low genetic divergence.  Indeed, bigdata
genomics recently investigated the supervised method, deep learning, for clustering the 1000 Genomes data by
ethnicity~\cite{Ferguson}.  \marginpar{``bigdata genomics'' seems to be a domain name \url{http://bdgenomics.org/} --
  how should we cite this}

In this paper we therefore systematically evaluate supervised and unsupervised methods for determining population
association and compare our results to those obtained using deep learning.
%\marginpar{TODO Rob can you think of a better description of what it does}
One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary
by fitting a linear combination of the feature variables.  In this appliction, it is not feasible to model interactions and
so the model is restricted to main effects.

%\marginpar{TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}.}  
We also investigate more complex supervised
approaches, namely Random Forests (RF), with the rationale being that interaction effects may dominate. A recent meta-study by
Locke {\it et al.}~\cite{Locke2015} found across 97 genomic loci associated with genetic susceptibility to obesity, only five
had an independent association signal and the rest seem to jointly predict this complex metabolic phenotype.  
We therefore investigate whether ethnicity is a similarly complex phenotype that requires higher order dependencies between
features to be taken into account. %, as done by RF and DL.

RF inherit many of the benefits of decision trees. It handles large numbers of variables transparently and
provides an unbiased estimate of the generalization error (in addition, it handles missing values and unbalanced
population class sizes). RF provide a measure of variable importance and allow for interactions between
variables to be investigated. However, in this problem, with many millions of variables, inspecting plots of interactions
is not feasible.  We note that decision trees (and random forests) model local variable interactions -- there are 
no main effects or global interaction effects.

@@

However, training a supervised learning method on such an expansive dataset as the 1000 Genomes Project phase 1 dataset is computationally 
challenging and traditional massively parallel approaches to data processing are not easily applicable to machine learning 
tasks, as they iteratively refine models based on information from the full dataset.

@@

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.  
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to a limitation of the Java virtual machine (maximum array size) that causes {\sc Mllib}'s
random forest implementation to fail. This error is due to the number of features present in each sample,
rather than the number of samples, and a potential solution is to either limit the features through feature selection, or to simply use a subset of the dataset.
We therefore use only chromosome 22 (494,328 variants) for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes Project dataset using the annotated labels of all
super-populations.  We perform 10-fold cross-validation and calculate the Adjusted Rand Index (ARI) for use as a metric for model quality.
This metric allows us to compare the classification results with \kMeans{} clustering results.
We also report the mean error for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.

We perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)). As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately,
with an average ARI over ten models of \(0.596 \pm 0.067\) (see table~\ref{RF}).
%Although capable of fitting a relatively accurate model (\(ARI=0.837\)), The average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\), however it is outperformed by the LR, with an ARI of \(0.923 \pm 0.015\).
This indicates that while applying a supervised learning method results in higher accuracy, a multivariate approach may not be needed to accurately fit ethnicity.

@@

\subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
In this section we investigate whether another multivariate approach, DL, is able to achieve a better result than
RF by again comparing it against \kMeans{} and logistic regression.  We compare against the reported accuracy of the
deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}.
As the reported results were performed on only three of the 14 populations, we similarly restrict our comparison to ASW
(Americans of African Ancestry in SW USA), CHB (Han Chinese in Beijing, China), and GBR (British in England and
Scotland).
